Supplemental Material for: The evolution of gene expression in the term placenta of viviparous mammals

### 2017-03-17 manuscript/manuscript.Rnw
use 13 eutherians instead of 14 of 14
include precise numbers for gene trees
 - Calculate permissive gene tree as well (FPKM ≥ 5)
 - Make sure to add back in the human name for humans even if there
   isn't a human species

<<load_core_placenta_tree_transcriptome,echo=FALSE,results="hide",cache.extra=file.info("../data/placenta_core_tree_transcriptome")[,"mtime"]>>=
load("../data/placenta_core_tree_transcriptome")
cpttl <-
    core.placenta.tree.transcriptome.long[!grepl("(min|median|max|all)_expression|all_but_",
                                                 Species) & housekeeping==FALSE,]
@

The
\Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if an ortholog pair of that gene was
expressed with $\text{FPKM} \ge 10$ in all eutherian species studied
(excluding the metatherian, \opossum) and was not a human housekeeping
gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}. Genes which
meet this criteria but are housekeeping genes are depicted in
Supplemental \cref{fig:core_placenta_transcriptome_housekeeping}.
\Cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently; for example, choosing an \ac{FPKM} cutoff
of 1 yields %
\Sexpr{cpt.distribution.shape[`Expression (FPKM)`>=1 & `Min. Exp. in`
  == '14 of 14 Species',max(num_genes)]} %
genes in the core transcriptome. %
In contrast to the shape of the number of genes included in the core
placenta transcriptome, the set of \human housekeeping genes expressed
all placentas is flatter (Supplemental
\cref{fig:core_housekeeping_transcriptome_shape}), and includes more
of the total fraction of genes (%
%
\Sexpr{core.placenta.transcriptome.long[housekeeping==TRUE,length(unique(human_name))]/
  core.placenta.transcriptome.long[,length(unique(human_name))]} %
%
%
vs %
\Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]/
  core.placenta.transcriptome.long[,length(unique(human_name))]}%
%
). An alternative approach to identify housekeeping genes is to use
the tissue specificity
index~\cite{Yanai.ea2005:Genome-widemidrangetranscriptionprofiles}.
Supplemental
\cref{tab:tissue_specificity_index_of_placenta_core_genes} lists the
tissue specificity index for all housekeeping and non-housekeeping
genes; housekeeping genes expressed in all placentas have a lesser
median tissue specificity index than non-housekeeping genes
(Supplemental \cref{fig:tissue_specificity_index_with_housekeeping}).
We also note that housekeeping genes have been defined in \human, and
may not be housekeeping genes in other species; generating tissue
expression atlases and examining the evolution of expression in
non-model organisms is a promising area of future research.

@@

If we use gene trees to define the core placenta transcriptome instead
of ortholog pairs (where the mean expression for genes in the tree
which are present must be $\text{FPKM} \ge 10$), we find
\Sexpr{cpttl[,length(unique(tree_genes))]} unique gene trees which
include nearly all of the genes present in the core placenta
transcriptome (Supplemental
\cref{fig:core_placenta_tree_transcriptome,fig:core_placenta_tree_transcriptome_housekeeping}).

>>>>

<<load_core_placenta_tree_transcriptome,echo=FALSE,results="hide",cache.extra=file.info(paste0("../data/placenta_core_tree_transcriptome",c("","_permissive")))[,"mtime"]>>=
load("../data/placenta_core_tree_transcriptome")
load("../data/placenta_core_tree_transcriptome_permissive")
cpttl <-
    core.placenta.tree.transcriptome.long[!grepl("(min|median|max|all)_expression|all_but_",
                                                 Species) & housekeeping==FALSE,]
@

The
\Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if an ortholog pair of that gene was
expressed with $\text{FPKM} \ge 10$ in all eutherian species studied
(excluding the metatherian, \opossum) and was not a human housekeeping
gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}. Genes which
meet this criteria but are housekeeping genes are depicted in
Supplemental \cref{fig:core_placenta_transcriptome_housekeeping}.
\Cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently; for example, choosing an \ac{FPKM} cutoff
of 1 yields %
\Sexpr{cpt.distribution.shape[`Expression (FPKM)`>=1 & `Min. Exp. in`
  == 'All 13 Eutherians',max(num_genes)]} %
genes in the core transcriptome. %
In contrast to the shape of the number of genes included in the core
placenta transcriptome, the set of \human housekeeping genes expressed
all placentas is flatter (Supplemental
\cref{fig:core_housekeeping_transcriptome_shape}), and includes more
of the total fraction of genes (%
%
\Sexpr{core.placenta.transcriptome.long[housekeeping==TRUE,length(unique(human_name))]/
  core.placenta.transcriptome.long[,length(unique(human_name))]} %
%
%
vs %
\Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]/
  core.placenta.transcriptome.long[,length(unique(human_name))]}%
%
). An alternative approach to identify housekeeping genes is to use
the tissue specificity
index~\cite{Yanai.ea2005:Genome-widemidrangetranscriptionprofiles}.
Supplemental
\cref{tab:tissue_specificity_index_of_placenta_core_genes} lists the
tissue specificity index for all housekeeping and non-housekeeping
genes; housekeeping genes expressed in all placentas have a lesser
median tissue specificity index than non-housekeeping genes
(Supplemental \cref{fig:tissue_specificity_index_with_housekeeping}).
We also note that housekeeping genes have been defined in \human, and
may not be housekeeping genes in other species; generating tissue
expression atlases and examining the evolution of expression in
non-model organisms is a promising area of future research.

@@

If we use gene trees to define the core placenta transcriptome instead
of ortholog pairs (where the mean expression for genes in the tree
which are present must be $\text{FPKM} \ge 10$), we find
\Sexpr{cpttl[,length(unique(tree_genes))]} unique gene trees which
include most
(\Sexpr{length(intersect(core.placenta.tree.human.genes.in.tree,
  core.placenta.transcriptome.long[,as.character(unique(human_name))]))/
  core.placenta.transcriptome.long[,length(unique(human_name))]*100}\%)
of the genes present in the core placenta transcriptome (Supplemental
\cref{fig:core_placenta_tree_transcriptome,fig:core_placenta_tree_transcriptome_housekeeping}).
Relaxing the mean gene tree expression requirement to
$\text{FPKM} \ge 5$ includes nearly all of the genes
(\Sexpr{length(intersect(core.placenta.tree.human.genes.in.permissive.tree,
  core.placenta.transcriptome.long[,as.character(unique(human_name))]))/
  core.placenta.transcriptome.long[,length(unique(human_name))]*100}\%).

### 2017-03-17 manuscript/supplemental_material.Rnw
include precise numbers for gene trees
 - Calculate permissive gene tree as well (FPKM ≥ 5)
 - Make sure to add back in the human name for humans even if there
   isn't a human species

\begin{figure}[tbh]
<<core_placenta_tree_transcriptome,fig.width=12,fig.height=8,out.width="\\textwidth",out.height="0.8\\textheight",echo=FALSE>>=
cpttl <-
    core.placenta.tree.transcriptome.long[!grepl("(min|median|max|all)_expression|all_but_",
                                                 Species) & housekeeping==FALSE,]
cpttl <- cpttl[,Species:=factor(gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species)),
                                levels=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",
                                            capfirst(names(species.names))))]
cpttl <- cpttl[,median_FPKM:=median(FPKM,na.rm=TRUE),by=tree_genes]
cpttl <- cpttl[,facet:="high"]
cpttl <- cpttl[median_FPKM < median(median_FPKM),facet:="low"]
print(ggplot(cpttl[FPKM > 0.1,],aes(x=reorder(tree_genes,median_FPKM),y=FPKM))+
      geom_boxplot(outlier.shape=NA)+
      scale_y_log10(breaks=c(10^1,10^2,10^3,10^4),
                    labels=trans_format("log10", math_format(10^.x)))+
      geom_point(aes(color=Species,shape=Species))+
      scale_shape_manual(NULL,##"Species",
                         guide=guide_legend(#nrow=1,
                                            keywidth=unit(3,"mm"),
                                            keyheight=unit(3,"mm"),
                                            label.theme=element_text(face="italic",
                                                                     angle=0,
                                                                     size=7),
                                            title.position="top"
                                            ),
                         values=0:18)+
      scale_color_discrete(NULL)+
      coord_flip()+
      xlab("Genes in Tree")+
      ylab("Gene Expression (FPKM)")+
      theme_bw() +
      expand_limits(y=c(10^4)) + 
      theme(text=element_text(size=7.5),
            legend.position=c(0.99,0.01),
            axis.text.y=element_text(face="italic",size=7.5),
            legend.margin=margin(0,0,0,0,unit="mm"),
            legend.spacing=unit(0,"mm"),
            legend.justification=c(1,0),
            plot.margin=unit(c(0,0,0,0),"mm"),
            strip.background = element_blank(),
            strip.text.x = element_blank()
            )
      + facet_wrap(~facet,scales="free")
      )
@ 
\caption[Tree-based Core Placenta Transcriptome]{
  Tree-based Core Placenta Transcriptome.
  \Sexpr{cpttl[,length(unique(tree_genes))]}
  gene trees with mean expression ($\mathrm{FPKM} \ge 10$) in the
  placenta of all \Sexpr{star.logs[,length(unique(species))]} species
  studied which are not housekeeping~%
  \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}
  ordered (from top to bottom) by median expression.}
\label{fig:core_placenta_tree_transcriptome}
\end{figure}

>>>>

\begin{figure}[tbh]
<<core_placenta_tree_transcriptome,fig.width=12,fig.height=8,out.width="\\textwidth",out.height="0.8\\textheight",echo=FALSE>>=
cpttl <-
    core.placenta.tree.transcriptome.long[!grepl("(min|median|max|all)_expression|all_but_",
                                                 Species) & housekeeping==FALSE,]
cpttl <- cpttl[!is.na(tree_genes),Species:=factor(gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species)),
                                levels=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",
                                            capfirst(names(species.names))))]
cpttl <- cpttl[,median_FPKM:=median(FPKM,na.rm=TRUE),by=tree_genes]
cpttl <- cpttl[,facet:="high"]
cpttl <- cpttl[median_FPKM < median(median_FPKM),facet:="low"]
print(ggplot(cpttl[FPKM > 0.1,],aes(x=reorder(tree_genes,median_FPKM),y=FPKM))+
      geom_boxplot(outlier.shape=NA)+
      scale_y_log10(breaks=c(10^1,10^2,10^3,10^4),
                    labels=trans_format("log10", math_format(10^.x)))+
      geom_point(aes(color=Species,shape=Species))+
      scale_shape_manual(NULL,##"Species",
                         guide=guide_legend(#nrow=1,
                                            keywidth=unit(3,"mm"),
                                            keyheight=unit(3,"mm"),
                                            label.theme=element_text(face="italic",
                                                                     angle=0,
                                                                     size=7),
                                            title.position="top"
                                            ),
                         values=0:18)+
      scale_color_discrete(NULL)+
      coord_flip()+
      xlab("Genes in Tree")+
      ylab("Gene Expression (FPKM)")+
      theme_bw() +
      expand_limits(y=c(10^4)) + 
      theme(text=element_text(size=7.5),
            legend.position=c(0.99,0.01),
            axis.text.y=element_text(face="italic",size=7.5),
            legend.margin=margin(0,0,0,0,unit="mm"),
            legend.spacing=unit(0,"mm"),
            legend.justification=c(1,0),
            plot.margin=unit(c(0,0,0,0),"mm"),
            strip.background = element_blank(),
            strip.text.x = element_blank()
            )
      + facet_wrap(~facet,scales="free")
      )
@ 
\caption[Tree-based Core Placenta Transcriptome]{
  Tree-based Core Placenta Transcriptome.
  \Sexpr{cpttl[,length(unique(tree_genes))]}
  gene trees with mean expression ($\mathrm{FPKM} \ge 10$) in the
  placenta of all \Sexpr{star.logs[,length(unique(species))]} species
  studied which are not housekeeping~%
  \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}
  ordered (from top to bottom) by median expression.}
\label{fig:core_placenta_tree_transcriptome}
\end{figure}
